---
title: "Installation and Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

To accomodate the vast majority of KI data scientists, the kitools package is available in both R and Python.

This installation and setup article covers both languages, but in the package usage articles, you can use the R/Python toggle in the navigation bar to change the code examples to the language of your choice.

## Sign up for Synapse

Before installing kitools, you should make sure you have a [Synapse](https://www.synapse.org) account. The [landing page of Synapse](https://www.synapse.org) contains straightforward directions for this. Make note of your Synapse email and password.

## Install kitools in Python

If you are using Python, you can install the kitools package using `pip`.

```{python}
pip install \
  --index-url https://test.pypi.org/simple/ \
  --extra-index-url https://pypi.org/simple \
  kitools
```

Note that kitools requires Python >= 3.5.

## Install kitools in R

In R, you can install the kitools package directly from Github using the `remotes` package:

```{r}
install.packages("remotes") # (if not already installed)
remotes::install_github("ki-tools/kitools-r")
```

The kitools R package is simply a wrapper around the Python package. This is done to keep the package functionality and stability in sync. We have built the R package in a way that you [need not be aware of anything Python-related](https://github.com/hafen/rminiconda). Python and the associated packages are installed for you as part of an initial configuration. To configure your environment (this only needs to be done once), run the following:

```{r}
library(kitools)
kitools_configure()
```

## Store Synapse credentials

One additional step for getting set up to use kitools is to register your Synapse credentials in R or Python so that they can be cached and you don't need to be prompted for them every time you do an operation that contacts Synapse.

In Python:

```{python}
import synapseclient
syn = synapseclient.Synapse()
syn.login(__email__, __password__, rememberMe = True)
```

In R:

```{r}
library(kitools)
syn_login(__email__, __password__)
```

## Using the package

Now that you are set up, take a look at ["KI Projects and Working With Data"](working-with-data.html).
